Tropical mathematics and the Lambda-calculus I

### 2023-07-31 CSL24/main.tex
Commit
commit

\author{Davide {Barbarossa}}{Dipartimento di Informatica, Universit\`a di Bologna, %[optional: Address], 
Italy% \and My second affiliation, Country 
\and \url{https://lipn.univ-paris13.fr/\~barbarossa/index.html} }{davide.barbarossa@unibo.it}{https://orcid.org/0000-0003-4608-8282}{%(Optional)Funding each author
}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\author{Paolo {Pistone}}{Dipartimento di Informatica, Universit\`a di Bologna, %[optional: Address], 
Italy% \and LIP, Universit\'e de Lyon, France 
\and \url{%WebPage
} }{paolo.pistone2@unibo.it}{%Orchid
}{%(Optional)Funding each author
}

@@

\funding{This work has been supported by the DIAPASoN project}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
We study the interpretation of the lambda-calculus in a framework based on tropical mathematics, by considering the relational semantics weighted on the tropical semiring. The aim is to provide a framework where both program metrics, based on the analysis of program sensitivity via Lipschitz conditions, and resource analysis, based on higher-order program differentiation, coexist. In order to do that, we consider a graded-types lambda-calculus and the differential lambda-calculus, both languages ensuring bounded duplications.
A typical result of such line of research is that, thanks to the tropical operations, a simply typed term is interpreted as a locally Lipschitz map, which is in turn decomposed as an inf of more and more sensitive Lipschitz ones, corresponding to the interpretations of the approximants given by the lambda-calculus Taylor expansion.
We finally describe some promising lines of research, showing in particular how such tropical model expresses natural optimisation problems of probabilistic calculi.
\end{abstract}

@@

\section{Tropical Semantics of Effectful Programs}\label{section22}
\input{section2bisbis}

@@

\section{Taylor Expansion and Lipschitz Approximation}\label{sec:TayLip}
\input{sectionTaylip}

>>>>

\author{Davide {Barbarossa}}{Dipartimento di Informatica - Scienza e Ingegneria, Universit\`a di Bologna, %[optional: Address], 
Italy% \and My second affiliation, Country 
\and \url{https://lipn.univ-paris13.fr/\~barbarossa/index.html} }{davide.barbarossa@unibo.it}{https://orcid.org/0000-0003-4608-8282}{%(Optional)Funding each author
}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\author{Paolo {Pistone}}{Dipartimento di Informatica - Scienza e Ingegneria, Universit\`a di Bologna, %[optional: Address], 
Italy% \and LIP, Universit\'e de Lyon, France 
\and \url{http://logica.uniroma3.it/pistone/
} }{paolo.pistone2@unibo.it}{https://orcid.org/0000-0003-4250-9051}{%(Optional)Funding each author
}

@@

\funding{This work has been supported by the ERC CoG 818616 DIAPASoN.}%optional, to capture a funding statement, which applies to all authors. Please enter author specific funding statements as fifth argument of the \author macro.

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
We study the interpretation of the lambda-calculus in a framework based on tropical mathematics, and we show that it provides a unified framework for both program  metrics, based on the analysis of program sensitivity via Lipschitz conditions, and resource analysis, based on higher-order program differentiation. 
To do that we focus on the semantics arising from the relational model weighted over the tropical semiring, and we discuss its application to the study of “best case” program behavior for languages with probabilistic and non-deterministic effects. Finally, we show that a general foundation for this approach is provided by an abstract correspondence between tropical algebra and Lawvere’s theory of generalized metric spaces.
%A typical result of such line of research is that, thanks to the tropical operations, a simply typed term is interpreted as a locally Lipschitz map, which is in turn decomposed as an inf of more and more sensitive Lipschitz ones, corresponding to the interpretations of the approximants given by the lambda-calculus Taylor expansion.
%We finally describe some promising lines of research, showing in particular how such tropical model expresses natural optimisation problems of probabilistic calculi.
\end{abstract}

@@

\section{Tropical Semantics and First Order Effectful Programs}\label{section22}
\input{section2bisbis}

@@

\section{Taylor Meets Lipschitz}\label{sec:TayLip}
\input{sectionTaylip}

### 2023-07-31 CSL24/section1.tex
commit

\subparagraph*{Two kinds of quantitative approaches}

Among such quantitative approaches, two different directions have received considerable attention.

@@

On the other hand, there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}. This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable for the probabilistic setting \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.

In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using inputs exactly once), plays a crucial role.
In metric semantics, linear programs correspond to \emph{non-expansive} maps, that is, to functions that do not increase distances; moreover, the possibility of duplicating inputs leads to interpret \emph{bounded} programs (i.e.~programs with a fixed duplication bound) as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider functions defined as \emph{power series}.

A natural question is thus whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow reconciled. At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
in metric models $n$ times duplication results in a \emph{linear} (hence Lipschitz) function $n\cdot x$, while in differential models this results in a \emph{polynomial} function $x^{n}$, hence not Lipschitz. The fundamental intuition of this work is then the observation that 
this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ precisely reads as the linear function $n\cdot x$.

@@

Coming back to our discussion on program semantics, tropical geometry might seem to provide precisely what look for, as it turns the monomials $x^{n}$ into the Lipschitz map  $n\cdot x$.
At this point, it is worth mentioning that a tropical variant of relational semantics has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail. Furthermore, connections between tropical linear algebra and metric spaces have also been observed \cite{Fuji} within the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}.
However, a thorough investigation of the interpretation of the $\lambda$-calculus within tropical mathematics has not yet been undertaken.

In this paper we make a first step in such direction, by demonstrating that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provide the desired bridge between differential and metric semantics, and we suggest how this approach might be used to study probabilistic and non-deterministic programs (this last aspect will be developed in more detail in a future paper of this series).

@@

\subparagraph*{This paper and our contributions}

@@

%In this paper we demonstrate that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provides the desired bridge between differential and metric semantics. Moreover, we show that the conceptual unification of these two approaches suggests ways in which techniques from resource-analysis could be used in sensitivity analysis and \emph{vice-versa}, paving the way for new  applications of tropical geometry to the  study of higher-order programs.

Our contributions in this paper are the following:
\begin{itemize}

\item after recalling the notions of tropical polynomials and power series, we show how these objects naturally arise in the \emph{best-case} analysis of probabilistic and non-deterministic programs, by studying a few simple examples. This is in Sections \ref{section2} and \ref{section22}.

\item We study the relational model over the tropical semiring, which provides a semantics of effectful extensions of $\STLC$ (simply typed $\lambda$-calculus) and $\mathrm{PCF}$. Notably, 
 we show that the functions interpreting higher-order programs, %which correspond to a 
generalizations of \emph{tropical power series} \cite{Porzio2021}, are locally Lipschitz-continuous and thus yield a full-scale metric semantics for the $\lambda$-calculus and its bounded fragments. This is in Sections \ref{section3} and~\ref{sec:tls}.

 \item We exploit the differential structure of the relational model to study the \emph{tropical Taylor expansion} of a $\lambda$-term, which can be seen as an approximation of the term by way of Lipschitz-continuous maps, and we show that it can be used to deduce approximated Lipschitz-constants for higher-order programs.
This is in Section \ref{sec:TayLip}.

@@

\item We conclude 
by putting the connection between the 
tropical, differential and metric viewpoints at the right level of generality.
We recall a well-known correspondence between Lawvere's \emph{generalized metric spaces} \cite{Lawvere1973, Stubbe2014} and modules over the tropical semi-ring \cite{Russo2007} and we show that it yields a model of the differential $\lambda$-calculus which extends the tropical relational model. This is in Section~\ref{sec:GMS}.
\end{itemize}
%
%\section{Bounded and Differential $\lambda$-Calculi}
%
%
%Bounded Simply Typed $\lambda$-calculus $\BSTLC$:
%$$
%A::= o \mid !_{n}A \multimap A
%$$
%
%
%Resource Simply Typed $\lambda$-calculus $\RSTLC$:
%$$
%A::= o \mid [A, \dots , A] \multimap A
%$$
%
%
%Define a translation of types $(-)^{\C R}$ from $\BSTLC$ to $\RSTLC$ by $o^{\C R}=o$ and $(!_{n}A\multimap B)^{\C R}=
%[\underbrace{A^{\C R},\dots, A^{\C R}}_{n\text{ times}}]\multimap B^{\C R}$.
%
%\begin{proposition}
%$\Gamma \vdash_{\BSTLC} M:A$ implies 
%$\Gamma^{\C R}\vdash_{\RSTLC}M:A^{\C R}$.
%\end{proposition}
%

>>>>

\subparagraph*{Two approaches to quantitative semantics}

Among such quantitative approaches, two have received considerable attention.

@@

On the other hand, there is the approach based on \emph{differential} \cite{difflambda} or \emph{resource-aware} \cite{Boudol1993} extensions of the $\lambda$-calculus, which is well-connected to the so-called \emph{relational semantics} \cite{Manzo2012, Manzo2013, dill} and has a syntactic counterpart in the study of \emph{non-idempotent} intersection types \cite{decarvalho2018, Mazza2016}. This family of approaches have been exploited to account for higher-order program differentiation \cite{difflambda}, to establish reasonable \emph{cost-models} for the $\lambda$-calculus \cite{Accattoli2021}, and have also been shown suitable to the probabilistic setting \cite{Manzo2013, Breuvart2018, PistoneLICS2022}.

In both approaches the notion of \emph{linearity}, in the sense of linear logic \cite{girardLl} (i.e.~of using inputs exactly once), plays a crucial role.
In metric semantics, linear programs correspond to \emph{non-expansive} maps, i.e.~to functions that do not increase distances, and the possibility of duplicating inputs leads to interpret programs with a fixed duplication bound as \emph{Lipschitz-continuous} maps \cite{Gaboardi2017}.
By contrast, in the standard semantics of the differential $\lambda$-calculus, linear programs correspond to linear maps, in the usual algebraic sense, while the possibility of duplicating inputs leads to consider functions defined as \emph{power series}.

A natural question, at this point, is whether these two apparently unrelated ways of interpreting linearity and duplication can be somehow reconciled. At a first glance, there seems to be a  ``logarithmic'' gap between the two approaches:
in metric models $n$ times duplication results in a {linear} (hence \emph{Lipschitz}) function $n x$, while in differential models this results in a {polynomial} function $x^{n}$, thus not Lipschitz. The fundamental intuition of this work is then the observation that 
this gap is naturally overcome once we interpret these functions in the framework of tropical mathematics, where, as we'll see, the monomial $x^{n}$ precisely reads as the linear function $n x$.

@@

Coming back to our discussion on program semantics, tropical mathematics seems to be just  what we look for, as it turns polynomial functions like $x^{n}$ into Lipschitz maps like $n x$.
At this point, it is worth mentioning that a tropical variant of the usual relational semantics of linear logic and the $\lambda$-calculus has already been considered \cite{Manzo2013}, and shown capable of capturing \emph{best-case} quantitative properties, but has not yet been studied in detail. Furthermore, connections between tropical linear algebra and metric spaces have also been observed \cite{Fuji} within the abstract setting of \emph{quantale-enriched} categories \cite{Hofmann2014, Stubbe2014}.
However, a thorough investigation of the interpretation of the $\lambda$-calculus within tropical mathematics has not yet been undertaken.

In this paper we make a first step in such direction, by demonstrating that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provide the desired bridge between differential and metric semantics, and we suggest how this approach might be used to study probabilistic and non-deterministic programs.% (an aspect that will be developed in more detail in a future paper of this series).

@@

\subparagraph*{Contributions and outline of the paper}

@@

%In this paper we demonstrate that the relational interpretation of the $\lambda$-calculus based on tropical mathematics does indeed provides the desired bridge between differential and metric semantics. Moreover, we show that the conceptual unification of these two approaches suggests ways in which techniques from resource-analysis could be used in sensitivity analysis and \emph{vice-versa}, paving the way for new  applications of tropical geometry to the  study of higher-order programs.
Our contributions in this paper are the following:
\begin{itemize}

\item We first show that tropical polynomials naturally arise in the {best-case} analysis of probabilistic and non-deterministic programs, turning the study of quantitative program behavior into a purely combinatorial problem. This is in Sections \ref{section2} and \ref{section22}.

\item We study the relational model over the tropical semiring, which provides a semantics of effectful extensions of $\STLC$ (simply typed $\lambda$-calculus) and $\mathrm{PCF}$. Notably, 
 we show that higher-order programs are interpreted by a  
generalizations of {tropical power series} \cite{Porzio2021}, and we show that these functions are locally Lipschitz-continuous, thus yielding a full-scale metric semantics. This is in Sections \ref{section3} and~\ref{sec:tls}.

 \item We exploit the differential structure of the relational model to study the \emph{tropical Taylor expansion} of a $\lambda$-term, which can be seen as an approximation of the term by way of Lipschitz-continuous maps, and we show that it can be used to compute approximated Lipschitz-constants for higher-order programs.
This is in Section \ref{sec:TayLip}.

@@

\item We conclude 
by putting the connection between the 
tropical, differential and metric viewpoints at the right level of abstraction and generality.
We recall a well-known correspondence between Lawvere's \emph{generalized metric spaces} \cite{Lawvere1973, Stubbe2014} and modules over the tropical semi-ring \cite{Russo2007} and we show that it yields a model of the differential $\lambda$-calculus which extends the tropical relational model. This is in Section~\ref{sec:GMS}.
\end{itemize}
%
%\section{Bounded and Differential $\lambda$-Calculi}
%
%
%Bounded Simply Typed $\lambda$-calculus $\BSTLC$:
%$$
%A::= o \mid !_{n}A \multimap A
%$$
%
%
%Resource Simply Typed $\lambda$-calculus $\RSTLC$:
%$$
%A::= o \mid [A, \dots , A] \multimap A
%$$
%
%
%Define a translation of types $(-)^{\C R}$ from $\BSTLC$ to $\RSTLC$ by $o^{\C R}=o$ and $(!_{n}A\multimap B)^{\C R}=
%[\underbrace{A^{\C R},\dots, A^{\C R}}_{n\text{ times}}]\multimap B^{\C R}$.
%
%\begin{proposition}
%$\Gamma \vdash_{\BSTLC} M:A$ implies 
%$\Gamma^{\C R}\vdash_{\RSTLC}M:A^{\C R}$.
%\end{proposition}
%

### 2023-07-31 CSL24/section2bis.tex
commit

\begin{definition}
A \emph{tropical power series}
%\footnote{In analogy with \cite{DanEhrh2011}, we could also call them \emph{tropical entire functions}.} 
%(tps) of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{j\in I}\{i_j x+c_{i_{j}}\}$, with $i_{j}\in\N$, $c_{i_{j}}\in\Lawv$.
in $k$-variables is a function $f:\Lawv^k\to\Lawv$ of shape $f(x)=\inf_{i\in I}\{ix + \matr f(i) \}$, where 
$I\subseteq\mathbb N^{k}$, $i x$ is the scalar product and $\matr f\in \Lawv^{{\mathbb N}^{k}}$ is a vector of coefficients.
When $I$ is finite, $f$ is called a \emph{tropical polynomial}. 
\end{definition}

In Section \ref{section3} we also consider tps in \emph{infinitely many} variables.

A tropical polynomial is always a piece-wise linear function since, e.g.\ in one variable, it has shape $f(x)=\min_{0\leq j\leq n}\{i_{j}x+c_{i_{j}}\}$
For example, the polynomials $\varphi_{n}(x)=\min_{0\leq j\leq n}\{jx+2^{-j}\}$
are illustrated in Fig.~\ref{fig:plot1} for $0\leq n \leq 4$.

@@

A \emph{tropical root} of a tropical power series $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable (i.e.~where the slope of $\varphi$ changes). When $\varphi$ is a polynomial, the roots of $\varphi$ coincide with the points where the minimum defining $\varphi$ is attained at least twice (see \autoref{fig:plot1}).
%For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $0\leq i \leq n$.
%Tropical roots yield the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as $f(x)=k\min\{x,x_{0}\}+ g(x)$.
Unlike in standard algebra, tropical roots of tropical polynomials can be computed in linear time \cite{Noferini2015}.

@@

Now, for a power series $f=\sum_{i\in I} a_i x^i$ (polynomials being the case for $I$ finite) with coefficients $a_i$ in a semiring $Q$, and for a fixed valuation $\mathrm{val}:Q\to \Lawv$, one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.
%Remark that tropicalising w.r.t.\ the trivial valuation gives $\trop^{\mathrm{val}_1} \sum_{i\in I} a_i x^i = \inf_{i\in I} \set{i\alpha} = \trop^{\mathrm{val}} \sum_{i\in I\subseteq\N} x^i$ for all valuation $\mathrm{val}$, because in $\Lawv$ the unit is $0$. %and the zero is $\infty$.
%Remember that a_i \neq 0 by def of formal power series
%That is, it corresponds to quotienting the coefficients in $Q$ for the idempotency of $+$.
\end{remark}

%For instance, the tps $\varphi(x,y)=\inf_{n}\{2x+(n+1)y,3x\}$ coincides with the polynomial $\min\{2x+y,3x\}$, since $2x+(n+1)y\geq 2x+y$ holds for all $n$. So the trivial valuation sends both the infinite series $\sum_{n}(x^{2}y^{n+1}+x^{3})$ and the polynomial $x^{2}y+x^{3}$ onto the same map.
Notice that in $\Lawv$ we have that $\preceq$ is $\geq$ and $\max^{\preceq}=\min$ and the zero is $\infty$.
Therefore, for instance, the tropicalisation (w.r.t.\ any valuation) sends the infinite power series $\sum_{n\geq 2} x^n$ to the tps $\varphi(x)=\inf_n (n+2)x$.
But the latter always coincides with the tropical polynomial $\varphi(x)=2x$, since $nx\geq 0$ for all $n\in\N$, which is in turn the tropicalisation of the polynomial $x^2$.
This shows that tropicalisation is \emph{not} in general an injective operation and in fact, as we show in \autoref{theorem:fepsilon}, tps (in finitely many variables) have a tendency to collapse, if not globally at least locally, onto tropical polynomials.

Similarly, by looking at Fig~\ref{fig:plot1} it appears that, \emph{far from $0$}, $\varphi$ behaves like some of the polynomials $\varphi_{n}$.
In particular, %for all $\epsilon >0$, $f$ coincides on $[\epsilon,\infty]$ with some polynomial $\varphi_{n}$. More, precisely, 
$\varphi$ coincides on $[\epsilon,\infty]$ with $\varphi_{n}$,
for $\epsilon \geq 2^{-(n+1)}$ (the smallest tropical root of $\varphi_{n}$).
However, at
%
%$x\in [\epsilon,\infty]$, with $\epsilon>0$
%It can be proven by hand that $\varphi(x)$ is a $\min$ for all $x>0$.
 $x=0$ we have that $\varphi(x=0)=\inf_{n\in\N} 2^{-n}=0$, and this is the only point where the $\inf$ is \emph{not} a $\min$.
Also, while the derivative of $f$ is bounded on all $(0,\infty)$, for $x\to 0^+$ it tends to $\infty$.
In fact, this is a general phenomenon, as showed below:
%shared by all tps with a finite number of variables, i.e.~of the form $f(x)=\inf_{n\in \mathbb N^{k}}\{nx + \matr f(n) \}$, where 
%$x\in \Lawv^{k}$, $n\in \mathbb N^{k}$, $nx$ is the scalar product and $\matr f\in \Lawv^{{\mathbb N}^{k}}$ is a vector of coefficients. 
% 
%%This phenomenon is reminiscent of \cite[Example 7]{Ehrhard2005},
%%%Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019
%%which actually motivated our first investigations.
%In fact, this behavior is shared by all tps with \emph{finitely} many variables, as shown by the following result (we identify $!\set{1,\dots,k}\simeq \N^k$, so the matrix of a tps $f$ with finitely many variables $x=x_1,\dots,x_k$ (and one output variable) is given as a $\matr f:\N^k\to\Lawv$, $f$ having shape $f(x)=\inf_{n\in \N^k}\set{nx+\matr f(n)}$, with $nx$ the scalar product).

\begin{theorem}\label{theorem:fepsilon}
For all tLs $f(x_{1},\dots, x_{k})=\inf_{n\in \mathbb N^{k}}\{n x+\matr f(n)\}$, for all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ such that 
% 
% \begin{enumerate}
%  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv^k$;
%  \item If $f(x_0)=\infty$ for some $x_0\in[0,\infty)^k$, then $\C F_\epsilon=\emptyset$;
  %\item 
$f$ coincides on all $[\epsilon,\infty]^k$ with $P_\epsilon( x):=\min_{n\in \C F_\epsilon}\set{n x+\matr f(n)}$.
% \end{enumerate}
\end{theorem}
\begin{proof}[Proof sketch]
Let $\C F_\epsilon$ be the set of $n\in\N^{k}$ such that 
$\widehat f(n)<\infty$ and $\widehat f(m)> \widehat f(n)+\epsilon$ holds for all $m\prec n$, where $\preceq$ is the pointwise order on $\N^k$.
The core of the proof is showing that this set is indeed finite and sufficient for computing $f$.
\end{proof}

As we'll see, the potential of transforming infinitary objects (i.e.~tps) into combinatorial ones (i.e.~tropical polynomials), is one of the most interesting features of tropical semantics.

>>>>

\begin{definition}
A \emph{tropical power series} (tps) 
%\footnote{In analogy with \cite{DanEhrh2011}, we could also call them \emph{tropical entire functions}.} 
%(tps) of one variable is a function $f:\Lawv\to\Lawv$ of shape $f(x)=\inf_{j\in I}\{i_j x+c_{i_{j}}\}$, with $i_{j}\in\N$, $c_{i_{j}}\in\Lawv$.
in $k$-variables is a function $f:\Lawv^k\to\Lawv$ of shape $f(x)=\inf_{i\in I}\{ix + \matr f(i) \}$, where 
$I\subseteq\mathbb N^{k}$, $i x$ is the scalar product and $\matr f\in \Lawv^{{\mathbb N}^{k}}$ is a vector of coefficients.
When $I$ is finite, $f$ is called a \emph{tropical polynomial}. 
\end{definition}
Hence, a unary tps is a function $f:\Lawv\to \Lawv$ of the form $f(x)=\inf_{i\in I}\{ix+a_{i}\}$, with $I\subseteq\mathbb N$ and the $a_{i}\in \Lawv$.
In Section \ref{section3} we also consider tps in \emph{infinitely many} variables.

A tropical polynomial is always a piece-wise linear function since, e.g.\ in one variable, it has shape $f(x)=\min_{0\leq j\leq n}\{i_{j}x+c_{i_{j}}\}$.
For example, the polynomials $\varphi_{n}(x)=\min_{0\leq j\leq n}\{jx+2^{-j}\}$
are illustrated in Fig.~\ref{fig:plot1} for $0\leq n \leq 4$.

@@

A \emph{tropical root} of a tps $\varphi$ is a point $x\in \Lawv$ where $\varphi$ is not differentiable (i.e.~where the slope of $\varphi$ changes). When $\varphi$ is a polynomial, the roots of $\varphi$ coincide with the points where the minimum defining $\varphi$ is attained at least twice (see \autoref{fig:plot1}).
%For instance, the tropical roots of $\varphi_{n+1}$ are of the form $2^{-(i+1)}$, for $0\leq i \leq n$.
%Tropical roots yield the usual factorization property of roots: if $x_{0}$ is a root of $f$, this factorizes as $f(x)=k\min\{x,x_{0}\}+ g(x)$.
Unlike in standard algebra, tropical roots of tropical polynomials can be computed in linear time \cite{Noferini2015}.

@@

Now, for a power series $f(x)=\sum_{i\in I} a_i x^i$ (polynomials being the case for $I$ finite) with coefficients $a_i$ in a semiring $Q$, and for a fixed valuation $\mathrm{val}:Q\to \Lawv$, one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.
%Remark that tropicalising w.r.t.\ the trivial valuation gives $\trop^{\mathrm{val}_1} \sum_{i\in I} a_i x^i = \inf_{i\in I} \set{i\alpha} = \trop^{\mathrm{val}} \sum_{i\in I\subseteq\N} x^i$ for all valuation $\mathrm{val}$, because in $\Lawv$ the unit is $0$. %and the zero is $\infty$.
%Remember that a_i \neq 0 by def of formal power series
%That is, it corresponds to quotienting the coefficients in $Q$ for the idempotency of $+$.
\end{remark}

%For instance, the tps $\varphi(x,y)=\inf_{n}\{2x+(n+1)y,3x\}$ coincides with the polynomial $\min\{2x+y,3x\}$, since $2x+(n+1)y\geq 2x+y$ holds for all $n$. So the trivial valuation sends both the infinite series $\sum_{n}(x^{2}y^{n+1}+x^{3})$ and the polynomial $x^{2}y+x^{3}$ onto the same map.
Notice that in $\Lawv$ we have that $\preceq$ is $\geq$ and $\max^{\preceq}=\min$ and the zero is $\infty$.
Therefore, for instance, the tropicalisation (w.r.t.\ any valuation) sends the infinite power series $\sum_{n\geq 2} x^n$ to the tps $\varphi(x)=\inf_n (n+2)x$.
But the latter always coincides with the tropical polynomial $\varphi(x)=2x$, since $nx\geq 0$ for all $n\in\N$, which is in turn the tropicalisation of the polynomial $x^2$.
This shows that tropicalisation is \emph{not} in general an injective operation and in fact, as we show in \autoref{theorem:fepsilon} below, tps (in finitely many variables) have a tendency to collapse, if not globally at least locally, onto tropical polynomials.

For instance, by looking at Fig~\ref{fig:plot1} it appears that, \emph{far from $0$}, $\varphi$ behaves like some of the polynomials $\varphi_{n}$.
In particular, %for all $\epsilon >0$, $f$ coincides on $[\epsilon,\infty]$ with some polynomial $\varphi_{n}$. More, precisely, 
$\varphi$ coincides on $[\epsilon,\infty]$ with $\varphi_{n}$,
for $\epsilon \geq 2^{-(n+1)}$ (the smallest tropical root of $\varphi_{n}$).
However, at
%
%$x\in [\epsilon,\infty]$, with $\epsilon>0$
%It can be proven by hand that $\varphi(x)$ is a $\min$ for all $x>0$.
 $x=0$ we have that $\varphi(x=0)=\inf_{n\in\N} 2^{-n}=0$, and this is the only point where the $\inf$ is \emph{not} a $\min$.
Also, while the derivative of $f$ is bounded on all $(0,\infty)$, for $x\to 0^+$ it tends to $\infty$.
In fact, this is a general phenomenon, as showed below:
%shared by all tps with a finite number of variables, i.e.~of the form $f(x)=\inf_{n\in \mathbb N^{k}}\{nx + \matr f(n) \}$, where 
%$x\in \Lawv^{k}$, $n\in \mathbb N^{k}$, $nx$ is the scalar product and $\matr f\in \Lawv^{{\mathbb N}^{k}}$ is a vector of coefficients. 
% 
%%This phenomenon is reminiscent of \cite[Example 7]{Ehrhard2005},
%%%Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019
%%which actually motivated our first investigations.
%In fact, this behavior is shared by all tps with \emph{finitely} many variables, as shown by the following result (we identify $!\set{1,\dots,k}\simeq \N^k$, so the matrix of a tps $f$ with finitely many variables $x=x_1,\dots,x_k$ (and one output variable) is given as a $\matr f:\N^k\to\Lawv$, $f$ having shape $f(x)=\inf_{n\in \N^k}\set{nx+\matr f(n)}$, with $nx$ the scalar product).

\begin{theorem}\label{theorem:fepsilon}
For all tps $f(x)=\inf_{n\in \mathbb N^{k}}\{n x+\matr f(n)\}$, for all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ such that 
% 
% \begin{enumerate}
%  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv^k$;
%  \item If $f(x_0)=\infty$ for some $x_0\in[0,\infty)^k$, then $\C F_\epsilon=\emptyset$;
  %\item 
$f$ coincides on all $[\epsilon,\infty]^k$ with $P_\epsilon( x):=\min_{n\in \C F_\epsilon}\set{n x+\matr f(n)}$.
% \end{enumerate}
\end{theorem}
\begin{proof}[Proof sketch]
Let $\C F_\epsilon$ be the set of $n\in\N^{k}$ such that 
$\widehat f(n)<\infty$ and $\widehat f(m)> \widehat f(n)+\epsilon$ holds for all $m\prec n$, where $\preceq$ is the pointwise order on $\N^k$.
The core of the proof is showing that this set is indeed finite and sufficient for computing $f$.
\end{proof}

As we'll see, the potential of collapsing infinitary objects (i.e.~tps) into combinatorial ones (i.e.~tropical polynomials), is one of the most intriguing features of tropical semantics.
